---
title: "Mapping CityBike Routes"
author: "Oleksiy Anokhin"
date: "July 18, 2017"
output:
  html_document:
    keep_md: true
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
---



## Setup

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, fig.width=10)


```


```{r}

library( dplyr )
library( ggplot2 )
library( ggmap )
library( leaflet ) 

```


# Steps to Visualize Biking Data Routes

The following steps are explained in this document.

1. Create a list of unique stations.  
2. Create a list of unique pairs of stations. 
3. Create route for each pair of stations. 
4. Create a map of NYC. 
5. Add all routes to the map. 


For this exercise we will use City Bike NYC data from January of 2015. It was obtained from the file posted on AWS, and saved as an RData file.



# Load Data


```{r}

# Data for Bike Trips

dat <- readRDS(gzcon(url("https://github.com/lecy/CityBikeNYC/raw/master/DATA/bikes.rds")))

str( dat ) # Check the structure - 'data.frame':	285552 obs. of  16 variables

names( dat )

# Data on Routes

```



# Step 1. Create a list of unique stations

```{r}



# Pick start station attributes

keep.these.vars <- c("start.station.id","start.station.name",
                     "start.station.latitude","start.station.longitude")

stations <- unique( dat[ keep.these.vars ])

nrow(stations) # 330 unique stations

names(stations) <- c("ID", "StationName", "LAT", "LON") # Rename column titles

stations <- stations[order(stations$ID),] # put station in ID order from 72 till 3002

rownames(stations) <- NULL

head(stations)




# The same with dplyr

# Select four columns from "dat" dataframe. 
test1 <- dat %>% 
  select( start.station.id, 
          start.station.name,
          start.station.latitude,
          start.station.longitude  )

# Identify unique stations 
# test2 <- dat %>% select( start.station.id,
#                          start.station.name,
#                          start.station.latitude, 
#                          start.station.longitude) %>% distinct
                        
test2 <- test1  %>% distinct                      


head( test2 )

nrow( test2 )

```


We have 330 unique stations in the city:


```{r}

# plot unique stations and their IDs on a map

```






# Step 2. Identify unique pairs of stations.

Index of all combinations


```{r}

# combinations of stations

x <- c("A","B","C")
t(combn(x, 2))


```


```{r}


# In a loop, something like this:

for( i in c("A","B","C") )
{
  
  for( j in c("A","B","C") )
  {
    
    paste( i, j, sep="-" )
    
  }
  
}
```





# Step 3. Create route for each pair of stations.

Routes are queried through the Google Maps API.

```{r eval=F}


for(i in 1:330)
{
  print(paste("LOOP NUMBER", i))
  flush.console()
  routes <- list()
  for(j in 1:330)
  {
    rt <- try(route(from=c(stations$LON[i], stations$LAT[i]), 
                      to=c(stations$LON[j], stations$LAT[j]), 
                      mode="bicycling",
                      structure="route" 
    ))
    route.name <- paste("S.", stations$ID[i], "_to_S.", stations$ID[j], sep="")
    rt <- cbind(rt, from.to=route.name)
    routes[[j]] <- rt
    names(routes)[j] <- route.name
    print(paste("I=", i, "J=", j))
    flush.console()
  }  # end of j loop
  
  id <- substr(1000 + i, 2, 4)
  list.name <- paste("RoutesFromStation", id, ".rda", sep="")
  save(routes, file=list.name)
}


```





# Step 4. Load a Map of NYC.


```{r}


# lots of ways to do this:

# qmap 
nyc <- qmap("New York City, NY", color='bw', zoom=13) 

# ggmpap
nyc_map <- get_map(location = c(lon = -74.00, lat = 40.71), maptype = "terrain", zoom = 11)
ggmap(nyc_map)


```

### Leaflet

```{r, eval=T}

# Leaflet 1
leaflet( dat ) %>% addTiles() %>% setView(-74.0059, 40.7128, zoom = 13)


# addTiles("http://korona.geog.uni-heidelberg.de/tiles/roadsg/x={x}&y={y}&z={z}", 
#                                           attribution='Map tiles by <a href="http://stamen.com">Stamen Design</a>, <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a> &mdash; Map data &copy; <a href="http://www.openstreetmap.org/copyright">OpenStreetMap</a>') 

```


```{r, eval=T}

# Leaflet 2
leaflet() %>%
  addTiles() %>%
  setView(-74.00, 40.71, zoom = 12) %>%
  addProviderTiles("CartoDB.Positron")


```



# Step 5. Add all routes to the map.

lots of ways to do this
and we will need to play with aesthetics

we can add multiple routes by grouping them,
or we can layer them using a loop






```{r, eval=F}


df <- do.call(rbind.data.frame, routes)

### create some fake trip data

unique.routes <- unique(df$from.to) # pick all unique routes 

route.frequency <- data.frame( route=unique.routes, 
                               trips=sample(1:50, length(unique.routes)))

head(route.frequency) 

df <- merge(df, route.frequency, by.x="from.to", by.y="route") # merge data

# Mapping data

nyc <- qmap("New York City, NY", color='bw', zoom=13)  

nyc + geom_path( aes( x=lon , y=lat, group=from.to), 
                   colour="#1E2B6A", data=df, alpha=0.5, size=df$trips/25)
```











```{css}
p {
color: black;
margin: 0 0 20px 0;
}

td {
    padding: 3px 10px 3px 10px;
    text-align: center;
}

table
{ 
    margin-left: auto;
    margin-right: auto;
    margin-top:80px;
    margin-bottom:100px;
}

h1, h2{
  margin-top:100px;
  margin-bottom:20px;
}

H5{
    text-align: center;
    color: gray;
    font-size:0.8em;
}

img {
    max-width: 90%;
    display: block;
    margin-right: auto;
    margin-left: auto;
    margin-top:30px;
    margin-bottom:20px;
}

pre {
  overflow-x: auto;
}

pre code {
   display: block; 
   padding: 0.5em;
   margin-bottom:20px;
}

code {
  font-size: 92%;
  border: 10px solid #F8F8F8;
  margin-bottom: 2px;
}

code[class] {
  background-color: #F8F8F8;
}

```



