---
title: "Build Routes Database"
author: "Jesse Lecy"
output:
  html_document:
    keep_md: true
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
---



## Setup

```{r setup, include=TRUE}

knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, fig.width=10)

library( ggmap )

```



# Example Route

```{r}


rt <- route( from=c(-73.96905,40.75002), 
             to=c(-73.99148,40.72229), 
             mode="bicycling",
             structure="route" )

rt


# plot the route

nyc <- qmap( "New York City, NY", color='bw', zoom=13 )  

nyc +  geom_path(  aes( x = rt$lon , y = rt$lat ), 
            colour="red", data=rt, alpha=1, size=2 )





```







# BUILD BIKE STATION DATABASE

In order to create a database of possible routes, we need to create a database with all bike stations.

To build the database let's omit trip info, include only station info, and apply the `unique()` function.

```{r}

dat <- readRDS(gzcon(url("https://github.com/lecy/CityBikeNYC/raw/master/DATA/bikes.rds")))

keep.these.vars <- c("start.station.id","start.station.name","start.station.latitude","start.station.longitude")

stations <- unique( dat[ keep.these.vars ] )

nrow( stations )

names( stations ) <- c("ID","StationName","LAT","LON")

stations <- stations[ order( stations$ID ) , ]

rownames( stations ) <- NULL

head( stations )


```








# BUILD A LIST OF ROUTES

The number of possible pairs of unique elements, or in other words the number of routes traveling from one station to another, can be calculated as follows:

* For all pairs including trips that start and end at the same station:  N * N

* If we omit trips that start and stop at the same station: N * (N-1)

* If we assume the route from A to B is the same as from B to A: ( N * (N-1) ) / 2

Since NYC contains a lot of one-way streets, let's not assume that routes are symmetric.

Note that the `route()` function in R returns a data frame object. We can store all possible routes as a list of data frames.

```{r}

routes <- list()


for( i in 1:5 )
{

   rt <- route( from=c(stations$LON[1], stations$LAT[1]), 
             to=c(stations$LON[i], stations$LAT[i]), 
             mode="bicycling",
             structure="route" )
   
   route.name <- paste( "S.", stations$ID[1], "_to_S.", stations$ID[i], sep="" )
   
   rt <- cbind( rt, from.to=route.name )

   routes[[i]] <- rt

   names(routes)[i] <- route.name

}



routes[ 1:3 ]

```


We can then find the specific route we need by referencing endpoints:


```{r}

route.name

names( routes )

routes[[ route.name ]]

```


Route names are built using the `paste()` function:

```{r}

paste( "S.", 72, "_to_S.", 79, sep="" )

# or more generally

paste( "S.", stations$ID[1], "_to_S.", stations$ID[2], sep="" )

```






# FULL LOOP THROUGH ALL ROUTES  

In order to find all possible pairs of stations (330*330=108,900 possible trips), we need to loop through all start points and all end points using a nested loop that will look something like this.

```{r, eval=F}

for( i in 1:330 )
{

   for( j in 1:330 )
   {

       from=i, to=j

   }

}

```


If you are creating large lists by querying expensive external APIs, I would recommend breaking the loops into parts and saving each list. This protects from a case where you lose internet connection or hit a query limit and the process stops without recording any progress.


You can save your progress by saving and loading lists as follows:

```{r, eval=FALSE}

saveRDS( routes, file="routes.rds " )

readRDS( "routes.rds" )

```



# EXCEEDING GGMAP DAILY LIMITS

The **ggmap** package utilizes the Google Maps API to generate data like geocoding and routes. Google places a 2,500 daily limit on free queries.

```{r}

ggmap_credentials()

routeQueryCheck()

```


To exceed this limit, you can create a premium account by registering through the developers page on Google and creating a paid account. It costs $0.50 for each 1,000 queries in excess of your free limit. Which means to build a database of 330 x 330 routes it will cost about $55.

https://developers.google.com/maps/web-services/




```{r}


ggmap_credentials()




API_KEY <- "AbCdEfGhIjKlMnOpQrStUvWxYz"    # fake key for demo

register_google( key=API_KEY, account_type="premium", day_limit=10000 )

ggmap_credentials()

routeQueryCheck()

```




# BUILD THE DATABASE


```{r, eval=F}



for( i in 1:330 )
{
  
  print( paste( "LOOP NUMBER", i ) )
  flush.console()
  
  routes <- list()
  
  for( j in 1:330 )
  {
    
     rt <- try( route( from=c(stations$LON[i], stations$LAT[i]), 
                  to=c(stations$LON[j], stations$LAT[j]), 
                  mode="bicycling",
                  structure="route" 
                 ) )
     
     route.name <- paste( "S.", stations$ID[i], "_to_S.", stations$ID[j], sep="" )
     
     rt <- cbind( rt, from.to=route.name )
  
     routes[[j]] <- rt
  
     names(routes)[j] <- route.name

     print( paste( "I=", i, "J=", j ) )
     flush.console()
     
  }  # end of j loop

  id <- substr( 1000 + i, 2, 4 )
  
  list.name <- paste( "RoutesFromStation", id, ".rda", sep="" )
  
  save( routes, file=list.name )
  
}








```


Once complete, combine all files into a single routes database.


```{r, eval=F}

file.names <- dir()

all.routes <- list()

for( i in 1:length(file.names) )
{

  load( file.names[i] )

  all.routes <- c( all.routes, routes )

}


saveRDS( all.routes, file="ALL_ROUTES_LIST.rds" )

```


# Fix Broken Routes

If your internet connection is bad, or if you hit query limits, you will have erroneous routes in your files. These can be fixed by identifying the errors and re-running those specific routes. Of the 108,000 routes I created, I had approximately 3,500 bad routes (error messages) due to bad internet connections.

```{r, eval=F}



table( unlist(lapply( all.routes, class )) )

route <- NULL
status <- NULL

for( i in 1:length(all.routes) )
{
    route[i] <- names( all.routes[i] )
    status[i] <- class( all.routes[[i]] )
}

success.log <- data.frame( route, status )

head( success.log )

nrow( success.log )

table( success.log$status )

these <- which( status == "try-error" )




route.names <- route[ these ]

route.names2 <- gsub( "S.", "", route.names )
from.to <- strsplit( route.names2, "_to_" )


for( i in 1:length(these) )
{
  
  print( paste( "ROUTE NAME:", route.names[i] ) )
  flush.console()

  from.id <- from.to[[i]][1]
  to.id <- from.to[[i]][2]

  from.station <- stations[ stations$ID == from.id , ]
  to.station <- stations[ stations$ID == to.id , ]

  rt <- try( route( from=c(from.station$LON, from.station$LAT), 
                  to=c(to.station$LON, to.station$LAT), 
                  mode="bicycling",
                  structure="route" 
                 ) )
     
     route.name <- route.names[i]

     if( names( all.routes[these[i]] ) == route.name )
     {
  
       all.routes[[these[i]]] <- rt

     } else( print("DOES NOT MATCH") )
  

  
}


table( unlist(lapply( all.routes, class )) )


# re-save

saveRDS( all.routes, file="ALL_ROUTES_LIST.rds" )


```






These can be translated into a data frame by:

```{r, eval=F}

# add route name to the data frames

for( i in 1:length(all.routes) )
{
    all.routes[[i]] <- cbind( all.routes[[i]], route=names(all.routes[i]) )
}

df <- do.call( rbind.data.frame, all.routes )

head( df )

saveRDS( all.routes, file="ALL_ROUTES_DF.rds" )

```



# Routes Available on GitHub

The routes database (list) and data frame have been posted to GitHub in the DATA repository. The can be loaded by:



```{r, eval=F}

rt <- readRDS( gzcon (url( "https://github.com/lecy/CityBikeNYC/raw/master/DATA/ALL_ROUTES_LIST.rds" ) ) )

rt <- readRDS( gzcon (url( "https://github.com/lecy/CityBikeNYC/raw/master/DATA/ALL_ROUTES_DF.rds" ) ) )

```








```{css}
p {
color: black;
margin: 0 0 20px 0;
}

td {
    padding: 3px 10px 3px 10px;
    text-align: center;
}

table
{ 
    margin-left: auto;
    margin-right: auto;
    margin-top:80px;
    margin-bottom:100px;
}

h1, h2{
  margin-top:100px;
  margin-bottom:20px;
}

H5{
    text-align: center;
    color: gray;
    font-size:0.8em;
}

img {
    max-width: 90%;
    display: block;
    margin-right: auto;
    margin-left: auto;
    margin-top:30px;
    margin-bottom:20px;
}

pre {
  overflow-x: auto;
}

pre code {
   display: block; 
   padding: 0.5em;
   margin-bottom:20px;
}

code {
  font-size: 92%;
  border: 10px solid #F8F8F8;
  margin-bottom: 2px;
}

code[class] {
  background-color: #F8F8F8;
}

```

